---
title: "Teach Yourself Shiny - Shiny tutorial by RStudio"
output: html_notebook
---

[source video](https://shiny.rstudio.com/tutorial/)

What Shiny can do: [Shiny Showcase](https://www.rstudio.com/products/shiny/shiny-user-showcase/)

# Part 1 - How to build a Shiny app

## Introduction

## R

Writing shiny app requires R knowledge

## App architecture

What user sees is only a front-end of an application, backend is server running R script everytime user interacts with an app.

App elements are manipulatable with drop-down menus, checkboxes, user input.

### App components

Each shiny app requires 2 component

- UI - webpage that user interacts with
- Server instructions

## App template

```{r eval=FALSE}
library(shiny)
ui <- fluidPage()

server <- function(input, output) {}

shinyApp(ui = ui, server = server)
```

Modern shiny app contains of 2 objects: 

- `ui` object that handles user interface for input and output
- `server` object that handles calculations

## Inputs and outputs

Build your app around `input()` and `output()` functions as parameters passed to  `ui` object

### `*Input()` functions

What `*Input()` function produces; given code:
```{r eval=F}
sliderInput(inputId = 'num',
            label = 'Choose a number',
            value = 25, min = 1, max = 100)
```
Produces html script
```{html}
<div class="container-fluid">
  <div class="form-group shiny-input-container">
    <label class="control-label" for="num">Choose a number</label>
    <input class="js-range-slider" id="num" data-min="1" data-max="100" data-from="25" data-step="1" data-grid="true" data-grid-num="9.9" data-grid-snap="false" data-prettify-separator="," data-prettify-enabled="true" data-keyboard="true" data-keyboard-step="1.01010101010101" data-data-type="number"/>
  </div>
</div>
```

Shiny provides about dozen input functions (available in help and documentation)

Standard `*Input()` function Syntax
```{r eval=F}
functionInput(<input_name>,<label>,<other_specific_parameters>)
```

### `*Output()` functions

Full list available in help

Standard `*Output()` function Syntax
```{r eval=F}
functionOutput(<output_name>,<other_specific_parameters>)
```

## The server function

It's being used to assemble inputs into outputs

You have to follow 3 rules to get reactivity:
- save objects to display to `output$`
- build objects to display with `render*()` functions
- use input values with `input$`
    - input values changes whenever a user changes the input

Reactivity 101 - reactivity automatically occurs whenever you use an input value to render an output object

## Sharing apps
- Shinyapps.io
- Shiny Server (community or Pro)

Every app has to contain `ui` and `server` objects in `app.R` file.
Older architecture (still compatible with Shiny Server) required puting `ui` and `server` parameters into separate files named accordingly (`ui.R`, `server.R`). This way of creating apps is not obsolete, it can still benefit when designing larger, more complex apps.

`app.R` script has to end with call to `shinyApp()` function

App and all it's dependencies (pictures, datasets, css, helper scripts etc.)
should be saved in one directory

# Part 2 - How to customize reactions

## Reactivity

App renders new output based on user input

## Reactive values

`input$num` - creates input value named "num"

`render*()` - builds reactive otput ito display in UI

```{r eval=F}
renderPlot(
  {
    # code in curly brackets will rebuild rendered object each time input value is changed
    hist(rnorm(input$num))
  }
)
```

Important note on scalability: When notified it is invalid, the object created by a `render*()` function will reurn the **entire block of code** associated with it


## Reactive functions

## `render*()`
- makes an object to display
- always save the result to `output$`
- `render*()` makes an observer object that has a **block of code** associated with it
- The object will rerun the entire doe block to update itself whenever it is invalidated

## Modularize code with `reactive()`

Make sure all objects using the same input values use the same data based on that input; call input once to generate data on the server side

To do this use `reactive()` function; `reactive()` builds a reactive object (reactive expression)

```{r eval=F}
data <- reactive({rnorm(input$num)})
```

Reactive expression 
- can be called like a function (eg. `data()`)
- it caches it value (avoid unnecesary computations) 

## Prevent reactions with `isolate()`

Prevent title field from updating the plot until user updates value of other, not isolated variable

It allows developer to create chain of reactivity

`isolate()` returns the result as a non-reactive value

```{r eval=F}
isolate({rnorm(input$num)})
```


## Trigger code with `observeEvent()` 

An action button - works as input function

```{r eval=F}
ui <- fluidPage(
  actionButton(inputId = "clicks",
    label = "Click me")
)
```

`observeEvent()` syntax
```{r eval=F}
observeEvent(
  # reactive values to respond to (observer invalidates ONLY when this value changes)
  input$clicks,{ 
    #code block to run whenever observer is invalidated
    print(input$clicks)
  }
)
```

Code wrapped in `observeEvent()` won't run even if it's valueas are being modified, until event is not observed (wchich triggers evaluation of the code)

`observe()` - also triggers code to run on server, uses same syntax as `render*()`, `reactive()` and `isolate()`

```{r eval=F}
observe({print(input$clicks)})
```

Observer will respond to every reactive value in the code and rerun it's code block

### Recap `observeEvent()`
- triggers code to run on the server
- specify precisely which reactive values should invalidate the observer
- use `observe()` for a more implicit syntax

## Delay reactions with eventReactive()

A reactive expression that only responds to specific values
```{r eval=F}
data <- eventReactive(
  # reactive value(s) to respond to 
  input$go, {
    # code used to build (and rebuild) object
    # this code will be treated as if it has beend isolated with isolate()
    ronorm(input$num)
  }
)
```

### Recap
- use `eventReactive()` to delay reactions
- `eventReactive()` creates a reactive expression
- you can specify precisely which reactive values should invalidate the expression

## Manage state with `reactiveValues()`

creates a list of reactive values to manipulate programmatically

```{r}
rv <- reactiveVAlues(
  # optional elements to add to the list
  data = rnorm(100)
)
```

### Recap 

- `reactiveValues()` creates a list of reactive values
- you can manipulate these values (usually with `observeEvent()`)



## Recap - Part 2

## Parting tips

# Part 3 - How to customize appearance

## Introduction

## Review of Parts 1 and 2

## HTML UI

## Adding static content

## Building layouts

## Panels and tabsets

## Prepackaged layouts

## CSS

## Recap - Part 3